---
title: "Using the 'bcgroundwater' R Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to bcgroundwater}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo=FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
```


You can install the 'bcgroundwater' R 📦 package directly from [GitHub](https://github.com/bcgov/bcgroundwater). To do so, you will 
need the [remotes](https://cran.r-project.org/web/packages/remotes/index.html) package:

```{r, eval=FALSE}
install.packages("remotes")
```

Next, install the `bcgroundwater` package using `remotes::install_github()`:

```{r, eval=FALSE}
library("remotes")
install_github("bcgov/bcgroundwater")
```

Then load the package:

```{r}
library(bcgroundwater)
```


## Get Your Data

First, decide which groundwater observation well is your well of interest from the British Columbia Provinicial Groundwater Observation Well Monitoring
Network [interactive map tool](https://www2.gov.bc.ca/gov/content?id=2D3DB9BA78DE4377AB4AE2DCEE1B409B). Here we will work with data from Groundwater Observation Well **#309**.
  
  
Download data on the well of interest&mdash;this may take a few moments.
  
  
If you are interested in daily averages, specify `which = "daily"`:

```{r, message=FALSE, warning=FALSE}
daily_data <- get_gwl(wells = 309, which = "daily")
head(daily_data)
```

If you are interested only in recent hourly data&mdash;within the last two years&mdash; specify `which = "recent"`:

```{r, message=FALSE, warning=FALSE}
recent_data <- get_gwl(wells = 309, which = "recent")
head(recent_data)
```

Otherwise, by default `which = "all"` which downloads all hourly data for that well:

```{r, message=FALSE, warning=FALSE}
data <- get_gwl(wells = 309)
head(data)
```



The Provincial [Groundwater Observation Well Network - Groundwater Levels Data](https://catalogue.data.gov.bc.ca/dataset/57c55f10-cf8e-40bb-aae0-2eff311f1685) is programatically sourced from the B.C. Data Catalogue, distributed under the [Open Government Licence - British Columbia](https://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61).


## Summarise & Visualise Your Data

Next, you can calculate the median monthly values:

```{r, message=FALSE, warning=FALSE}
monthly_data <- monthly_values(data)
head(monthly_data)
```

You can plot the seasonal patterns in the water levels of the well with
`gwl_monthly_plot()`. This displays the mean deviation from the yearly average,
as well as the range of variation:

```{r, message=FALSE, warning=FALSE, fig.align='center'}
monthlyplot <- gwl_monthly_plot(monthly_data, last12 = TRUE)
plot(monthlyplot)
```

To perform the trend analysis, you will need to generate a full regular time series
with no gaps in the dates. `make_well_ts()` does this for you, removing consecutive missing values near the
beginning or the end of the time series&mdash;due to the strong influence of
these regions on trend analysis&mdash;and interpolating the remaining missing values:

```{r, fig.width=6, fig.height=4, message=FALSE, warning=FALSE}
full_monthly_data <- make_well_ts(monthly_data)
head(full_monthly_data)
```

Ten years is often considered the minimum time series length for meaningful long-term trend analysis, and it is useful to also consider an acceptable allowable "maximum" amount of missing (interpolated) data. You can use the `dplyr` package to generate some summary information about the well dataset to help you decide if it is approrpiate to conduct a trend analysis:

```{r, message=FALSE, warning=FALSE}
library(dplyr)

time_series_attr <- full_monthly_data %>%
  group_by(Well_Num) %>%
  summarise(dataStart = as.Date(min(Date)), 
            dataEnd = as.Date(max(Date)), 
            dataYears = as.numeric(dataEnd - dataStart) / 365, 
            nObs = n(), 
            nMissing = length(med_GWL[nReadings == 0]), 
            percent_missing = round(nMissing/nObs*100, 1)) %>% 
  select(Well_Num, dataYears, nMissing, percent_missing) 
time_series_attr
```

Groundwater Observation Well **#309** has over 28 years of consecutive water level data with minimal missing data (2%), ideal for long-term trend analysis.

For trend analysis over a long time series, it is often beneficial to test for
trends with yearly averages, otherwise serial autocorrelation can be a problem
(even with pre-whitening). These can be calculated using the `dplyr`
package:

```{r, message=FALSE, warning=FALSE}
library(dplyr)

annual_data <- full_monthly_data %>%
  select(-yearmonth) %>%
  group_by(EMS_ID, Well_Num, Year) %>%
  summarize(nReadings = n(),
            mean_GWL = mean(med_GWL),
            SD = sd(med_GWL),
            med_GWL = median(med_GWL),
            q95_GWL = quantile(med_GWL, 0.95),
            n_months = n()) %>% 
  filter(n_months == 12)
head(annual_data)
```

## Analyse & Visualise Your Data

You can now calculate the long-term trend:

```{r, message=FALSE, warning=FALSE}
trends <- gwl_zyp_test(annual_data, byID = "Well_Num", col = "mean_GWL")
trends
```


We we will use the results from the yuepilon method, and categorize the observed long-term trend using the combined statistical & environmental significance criteria developed for the [Environmental Reporting BC indicator on trends in groundwater levels](http://www.env.gov.bc.ca/soe/indicators/water/groundwater-levels.html): Stable, Increasing, Moderate Rate of Decline or Large Rate of Decline:

```{r, message=FALSE, warning=FALSE}
result <-  trends %>% 
  filter(test_type == "yuepilon") %>%
  mutate(trend_category = case_when(trend >= 0.1 & sig < 0.05 ~ "Large Rate of Decline",
                           trend >= 0.03 & trend < 0.1 & sig < 0.05 ~ "Moderate Rate of Decline",
                           trend <= -0.03 & sig < 0.05 ~ "Increasing",
                           TRUE ~ "Stable")) 
result
```


Finally, you can plot the time series and the significant trend lines, optionally showing any interpolated values. Showing the overlaid trend line when the trend category is stable is also optional:

```{r, message=FALSE, warning=FALSE, fig.align='center'}
trend_plot <- gwl_area_plot(full_monthly_data, trend = result$trend,
                          intercept = result$intercept, sig = result$sig, 
                          trend_category = result$trend_category,
                          mkperiod = "annual", showInterpolated = TRUE,
                          show_stable_line = TRUE)
plot(trend_plot)
```


## Getting Help

Type `?bcgroundwater` in the R console for full documentation on the package functions and arguments.

Details, rationales and references on the trend and other analytical methods used under the hood 🚘 in the 'bcgroundwater' 📦 can be found here: <http://www.env.gov.bc.ca/soe/indicators/water/groundwater-levels.html>. 

`r Sys.Date()`







