---
title: "Using the 'bcgroundwater' R Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bcgroundwater}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo=FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
```


You can install the 'bcgroundwater' R `r emo::ji("package")` directly from [GitHub](https://github.com/bcgov/bcgroundwater). To do so, you will 
need the [remotes](https://cran.r-project.org/web/packages/remotes/index.html) package:

```{r, eval=FALSE}
install.packages("remotes")
```

Next, install the `bcgroundwater` package using `remotes::install_github()`:

```{r, eval=FALSE}
library("remotes")
install_github("bcgov/bcgroundwater")
```

Then load the package:

```{r}
library(bcgroundwater)
```


## Get Your Data

First, decide which groundwater observation well is your well of interest from the British Columbia Provinicial Groundwater Observation Well Monitoring
Network [interactive map tool](https://www2.gov.bc.ca/gov/content?id=2D3DB9BA78DE4377AB4AE2DCEE1B409B). Here we will work with data from Groundwater Observation Well **#309**.
  
  
Download data on the well of interest&mdash;this may take a few moments.
  
  
If you are interested in daily averages, specify `which = "daily"`:

```{r, message=FALSE, warning=FALSE}
daily_data <- get_gwl(wells = 309, which = "daily")
head(daily_data)
```

If you are interested only in recent hourly data&mdash;within the last two years&mdash; specify `which = "recent"`:

```{r, message=FALSE, warning=FALSE}
recent_data <- get_gwl(wells = 309, which = "recent")
head(recent_data)
```

Otherwise, by default `which = "all"` which downloads all hourly data for that well:

```{r, message=FALSE, warning=FALSE}
data <- get_gwl(wells = 309)
head(data)
```



The Provincial Groundwater Observation Well Network - Groundwater Levels Data is programatically sourced from the B.C. Data Catalogue, distributed under the [Open Government Licence - British Columbia](https://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61):  <https://catalogue.data.gov.bc.ca/dataset/57c55f10-cf8e-40bb-aae0-2eff311f1685>


## Tidy & Visualise Your Data

Next, calculate the median monthly values:

```{r, message=FALSE, warning=FALSE}
monthly_data <- monthlyValues(data)
head(monthly_data)
```

You can plot the seasonal patterns in the water levels of the well with
`gwlMonthlyPlot()`. This displays the mean deviation from the yearly average,
as well as the range of variation:

```{r, message=FALSE, warning=FALSE}
monthlyplot <- gwlMonthlyPlot(monthly_data, last12 = TRUE)
plot(monthlyplot)
```

To perform the trend analysis, you will need to generate a full regular time series
with no gaps in the dates. `makeWellTS()` does this for you&mdash;removing consecutive missing values near the
beginning or the end of the time series, due to the strong influence of
these regions on trend analysis, and interpolating the remaining missing values:

```{r, fig.width=6, fig.height=4, message=FALSE, warning=FALSE}
full_monthly_data <- makeWellTS(monthly_data)
head(monthly_data)
```

Ten years is often considered the minimum time series length for meaningful long-term trend analysis, and it is useful to also consider an acceptable allowable "maximum" amount of missing (interpolated) data. Let's use the `dplyr` package to generate some summary data about the well data:

```{r, message=FALSE, warning=FALSE}
library(dplyr)

time_series_attr <- full_monthly_data %>%
  group_by(Well_Num) %>%
  summarise(dataStart = as.Date(min(Date)), 
            dataEnd = as.Date(max(Date)), 
            dataYears = as.numeric(dataEnd - dataStart) / 365, 
            nObs = n(), 
            nMissing = length(med_GWL[nReadings == 0]), 
            percent_missing = round(nMissing/nObs*100, 1)) %>% 
  select(Well_Num, dataYears, nMissing, percent_missing) %>% 
  print()
```

Groundwater Observation Well **#309** has over 28 years of consecutive water level data with minimal missing data (2%), ideal for long-term trend analysis.

For trend analysis over a long time series, it is often beneficial to test for
trends with yearly averages, otherwise serial autocorrelation can be a problem
(even with pre-whitening). These can be calculated easily using the `dplyr`
package:

```{r, message=FALSE, warning=FALSE}
library(dplyr)

annual_data <- full_monthly_data %>%
  select(-yearmonth) %>%
  group_by(EMS_ID, Well_Num, Year) %>%
  summarize(nReadings = n(),
            mean_GWL = mean(med_GWL),
            SD = sd(med_GWL),
            med_GWL = median(med_GWL),
            q95_GWL = quantile(med_GWL, 0.95))
```

## Analyse & Visualise Your Data

You can now calculate the long-term trend:

```{r, message=FALSE, warning=FALSE}
trends <- gwlZypTest(annual_data, byID = "Well_Num", col = "mean_GWL")
trends
```

Finally, plot the time series with the trend overlaid (we will use the results
from the yuepilon method), optionally with interpolated values overlaid:

```{r, message=FALSE, warning=FALSE}
trend_plot <- gwlAreaPlot(full_monthly_data, trend = trends$trend[1],
                          intercept = trends$intercept[1], sig = trends$sig[1],
                          state = "Stable", mkperiod = "annual", showInterpolated = TRUE)
plot(trend_plot)

```

## Getting Help

Type `?bcgroundwater` in the R console for full documentation on the package functions and arguments.

Details, rationales and references on the trend and other analytical methods used under the hood `r emo::ji("oncoming_automobile")` in the 'bcgroundwater' `r emo::ji("package")` can be found here: <http://www.env.gov.bc.ca/soe/indicators/water/groundwater-levels.html>. 

`r Sys.Date()`







