---
title: "Using the bcgroundwater package"
date: "`r Sys.Date()`"
output_format: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the bcgroundwater package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r set-options, echo=FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
```

First, download some data for a well of interest from the B.C. Observation Well
Network [interactive map tool](http://www.env.gov.bc.ca/wsd/data_searches/obswell/map/obsWells.html). 
We have downloaded the data from Observation Well 309 and saved it.

Load the package, and read in the data:

```{r}
library("bcgroundwater")
data <- readGWLdata("gwl_report.csv", emsID = "E208036")
```

The data will be in the following format: 

```{r echo = FALSE}
head(data)
```

Next, calculate the median monthly values:

```{r}
monthly_data <- monthlyValues(data)
head(monthly_data)
```

You can plot the seasonal patterns in the water levels of the well with
`gwlMonthlyPlot()`:

```{r}
monthlyplot <- gwlMonthlyPlot(monthly_data, last12 = TRUE)
plot(monthlyplot)
```

To perform the analysis, you will need to generate a full time series with no
gaps. `makeWellTS()` does this for you, interpolating the missing values:

```{r, fig.width=6, fig.height=4}
full_monthly_data <- makeWellTS(monthly_data)
head(monthly_data)
```

For trend analysis over a long time series, it is often beneficial to test for 
trends with yearly averages, otherwise serial autocorrelation can be a problem 
(even with pre-whitening). These can be calculated easily using the `dplyr`
package:

```{r}
library("dplyr")

annual_data <- full_monthly_data %>%
  select(-yearmonth) %>% 
  group_by(EMS_ID, Well_Num, Year) %>%
  summarize(nReadings = n()
            , mean_GWL = mean(med_GWL)
            , SD = sd(med_GWL)
            , med_GWL = median(med_GWL)
            , q95_GWL = quantile(med_GWL, 0.95)) %>% 
  as.data.frame(stringsAsFactors = FALSE)
```

You can now calculate the trend:

```{r}
trends <- gwlZypTest(annual_data, col = "mean_GWL")
trends
```

Finally, plot the time series with the trend overlaid (we will use the results
from the yuepilon method), optionally with interpolated values overlaid:

```{r, warning=FALSE}
trend_plot <- gwlAreaPlot(full_monthly_data, trend = trends$trend[1], 
                          intercept = trends$intercept[1], sig = trends$sig[1], 
                          state = "Stable", mkperiod = "annual", showInterpolated = TRUE)
plot(trend_plot)

```

